---
ttitle: "Analysis 3: Fecundity analysis"
author: "Joe Gunn"
date: "2024-09-19"
output: html_document
---

# <b>Project:</b> Assessing phenotypic response and transgenerational inheritance following insecticide exposure in Colorado Potato Beetles (<i>Leptinotarsa decemlineata</i>; CPB) 
<font size="+1"> We assessed inter- and trans-generational effects of sublethal insecticide exposure on four larval performance phenotypes (fecundity, herbivory, mobility, and development) in Colorado potato beetles (<i>Leptinotarsa decemlineata</i>; CPB). We implemented a replicated inbred full-sibling pedigree design, beginning with 12 "founder" mated pair lineages and continuing for four successive generations (full siblings within lineages were inbred in each generation). For the transgenerational experiment, F<sub>0</sub> larvae (offspring of founder pairs) were divided into two "background" groups (exposed, i.e., exposed to a sub-lethal [10-ppm] dose of the neonicotinoid insecticide imidacloprid; and control, i.e., exposed to an equivalent dose of molecular-grade water). All mated pair beetles across lineages, treatments, and generations were monitored for daily fecundity. A subset of third instar larvae were collected from each treatment within each lineage for each generation and monitored for three  performance traits, including larval movement (mobility), feeding rate (herbivory), and time to discrete life stages (development). Performance traits were monitored in three successive generations (F<sub>1</sub>-F<sub>3</sub>) without treatment exposure and compared to the F<sub>0</sub> generation to test for inter- and transgenerationally-inherited "memory" of phenotypic variation. An additional subset of larvae from each generation were re-exposed to the treatments, both experimental (insecticide-exposed) and control (water), to test for potential inheritance of a phenotypic "response" to insecticide stress.</font>

# <b>Specific Aim:</b> Fecundity analysis
<font size="+1">In this analysis, we assess differences in fecundity (i.e., average daily number of eggs oviposited) among mated pairs in the transgenerational experiment by generation (i.e., F0, F1, F2, F3), lineage (i.e., L1, L2, ... L10), and treatment (i.e., control and exposed). We use average daily egg counts over the course of adult lifespan as our metric of fecundity rather than total number of eggs laid, because some mated pairs survived longer than other pairs and therefore had the opportunity to lay a greater number of eggs. </font>

## Phases of Analysis
### Phase 1: Data summary statistics
### Phase 2: Fecundity analysis

### Libraries needed for analysis
```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(writexl)
library(rstatix)
library(lme4)
library(rsq)
library(patchwork)
library(forcats)
library(car)
```

## PHASE 1: DATA SUMMARY STATISTICS
In this phase of the analysis, we calculate mean, standard deviation, and variance for average fecundity within experimental groups of interest. 

### STEP 1: Calculate summary statistics for fecundity data; run the Rmd chunk below.
In this step, we calculate mean, standard deviation, and variance for average fecundity (i.e., average number of eggs per day across lineages) for all experimental groups.

##### Calculate summary statistics for fecundity data:
```{r}
# load average fecundity data
load("data/average_fecundity.Rda")

# get sample size (n), mean, standard deviation (sd), and standard error (se) for average fecundity by generation and treatment
average_fecundity %>%
  group_by(adult_generation, background) %>% 
  summarize(n = n(),
            mean = mean(average_number_eggs),
            sd = sd(average_number_eggs),
            se = sd/sqrt(n()))
```

### STEP 2: Test for correlation between adult mass and fecundity.
In this step, we determine whether they is a significant correlation between adult mass and fecundity (i.e., number of eggs laid) to see if we may need to account for mass in our model of fecundity. We might expect mass to be positively correlated with number of eggs due to greater potential for allocation of resources to reproduction in larger individuals. If a significant correlation exists between adult mass and average daily fecundity, we will need to account for it in our downstream models. First, we determine whether mass data are fitted to a normal distribution. Next, we assess the correlation between mass and mean daily fecundity based on the appropriate underlying distribution.

#### 2a: Test fit of mass and fecundity data to a normal distribution.

##### 2a.1. Test fit of mass and fecundity data to a normal distribution; run the Rmd chunk below.

##### Test fit of mass and egg data to normal distribution:
```{r}
# load fecundity data
load("data/fecundity.Rda")

# get female (F) and male (M) mass data alongside mean number of eggs for each mated pair
fecundity <- fecundity %>%
  group_by(pair_id, background, `F`, M) %>%
  summarize(mean = mean(number_eggs))

# get control data only
c <- fecundity %>%
  filter(background == "C")

# get exposed data only
e <- fecundity %>%
  filter(background == "E")

# Test fit of mass data to normal distribution
shapiro.test(c$`F`)
shapiro.test(c$M)
shapiro.test(e$`F`)
shapiro.test(e$M)

# Test fit of egg data to normal distribution
shapiro.test(c$mean)
shapiro.test(e$mean)
```
<b>Analysis results:</b> <br>

<b>Shapiro-Wilk p-values for male and female mass in each background:</b><br>
F_C: p = 0.1063 <br>
F_E: p = 0.4594 <br>
M_C: p = 0.8849 <br>
M_E: p = 0.4455 <br>

<b>Shapiro-Wilk p-values for average daily fecundity in each background:</b><br>
Mean_C: p = 0.2947 <br>
Mean_E: p = 0.05107 <br>

For both female and male mass and mean number of eggs, neither background group significantly deviated from a normal distribution, so we assumed a normal distribution for linear mixed effect models to test for correlations.  

##### 2a.2. Plot empirical distributions for mass and fecundity data; run the Rmd chunk below.

##### Plot empirical data distribution for mass and fecundity: 1) `figures/mass_distributions.jpeg`; 2) `figures/fecundity_distribution.jpeg`
```{r}
# load fecundity data
load("data/fecundity.Rda")

# get female (F) and male (M) mass data alongside mean number of eggs for each mated pair
fecundity <- fecundity %>%
  group_by(pair_id, background, `F`, M) %>%
  summarize(mean = mean(number_eggs))

# plot female mass distribution
f_plot <- ggplot(fecundity, aes(x = `F`)) +
  geom_histogram(binwidth = 10, color = "black", fill = "grey", linewidth = 1) +
  theme_set(theme_cowplot(12)) +
  scale_x_continuous(breaks = seq(from = 0, to = 180, by = 30), limits = c(0,180)) +
  scale_y_continuous(breaks = seq(from = 0, to = 15, by = 3), limits = c(0,15)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 30, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  xlab("Mass (mg)") +
  ylab("Frequency") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot female mass distribution
m_plot <- ggplot(fecundity, aes(x = M)) +
  geom_histogram(binwidth = 10, color = "black", fill = "grey", linewidth = 1) +
  theme_set(theme_cowplot(12)) +
  scale_x_continuous(breaks = seq(from = 0, to = 180, by = 30), limits = c(0,180)) +
  scale_y_continuous(breaks = seq(from = 0, to = 15, by = 3), limits = c(0,15)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 30, hjust = 0.5)) + 
  theme(axis.title.y = element_blank()) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  xlab("Mass (mg)") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot mass distributions together
jpeg("figures/mass_distributions.jpeg", width = 10000, height = 4000, res = 700)

f_plot + m_plot + plot_layout(nrow = 1, ncol = 2)

dev.off()

# plot mean fecundity distribution
jpeg("figures/fecundity_distribution.jpeg", width = 5000, height = 4000, res = 500)

ggplot(fecundity, aes(x = mean)) +
  geom_histogram(binwidth = 5, color = "black", fill = "grey", linewidth = 1) +
  theme_set(theme_cowplot(12)) +
  scale_x_continuous(breaks = seq(from = 0, to = 60, by = 10), limits = c(0,60)) +
  scale_y_continuous(breaks = seq(from = 0, to = 15, by = 3), limits = c(0,15)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 30, hjust = 0.5)) + 
  theme(axis.title.y = element_blank()) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  xlab("Mean fecundity") +
  ylab("Frequency") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

dev.off()
```

#### 2b: Test correlation of mass and fecundity.

##### 2b.1. Test correlations between male and female mass and average daily fecundity; run the Rmd chunk below.

##### Assess correlations between male and female mass and fecundity: 
```{r}
# load fecundity data
load("data/fecundity.Rda")

# get female (F) and male (M) mass data alongside mean number of eggs for each mated pair
fecundity <- fecundity %>%
  group_by(pair_id, background, `F`, M) %>%
  summarize(mean = mean(number_eggs))

#test for correlation between mean fecundity and male mass
f_mean <- lmer(mean ~ `F` + (1|background),
               data = fecundity)

# test for correlation between mean fecundity and male mass
m_mean <- lmer(mean ~ M + (1|background),
               data = fecundity)

# summarize model results for fecundity and female mass
anova(f_mean) # get F values
Anova(f_mean) # get p-values
rsq.lmm(f_mean) # get r-squared values

# summarize model results for fecundity and male mass
anova(m_mean) # get F values
Anova(m_mean) # get p-values
rsq.lmm(m_mean) # get r-squared values
```

<b>Analysis results:</b> <br>

<b>Coefficients of linear mixed effect models for pairwise metric comparisons:</b><br>
<b>fecundity and female mass: F = 2.027; p = 0.1545; r-squared = 0.03069904 </b><br>
<b>fecundity and male mass: F = 0.1944; p = 0.6593; r-squared = 0.003027575 </b><br>

Given that these metrics are not correlated, we opted to omit mass from all downstream model analyses.

## PHASE 2: FECUNDITY ANALYSIS
In this phase of the analysis, we assess differences in fecundity (i.e., average number of eggs laid per day) background exposure statuses across generations, accounting for variation in lineages as a random effect in a general linear mixed-effect model framework. 

##### Plot daily fecundity: `figures/daily_fecundity.pdf` 
```{r}
# load fecundity data
load("data/fecundity.Rda")

# group data by generation, background, and days since oviposition and summarize mean and se of egg number (fecundity) across lineages. Calculate global and background-specific means for plotting.
fecundity <- fecundity %>%
  select(lineage:background, number_eggs, days_since_oviposition) %>%
  group_by(adult_generation, background, days_since_oviposition) %>%
  summarize(n = n(),
            mean = mean(number_eggs),
            sd = sd(number_eggs),
            se = sd/sqrt(n)) %>%
  group_by(adult_generation) %>%
  mutate(global_mean = mean(mean)) %>%
  group_by(adult_generation, background) %>%
  mutate(background_mean = mean(mean))

# isolate only F0 data
f0 <- fecundity %>%
  filter(adult_generation == "F0")

# isolate only F1 data
f1 <- fecundity %>%
  filter(adult_generation == "F1")

# isolate only F2 data
f2 <- fecundity %>%
  filter(adult_generation == "F2")

# isolate only F3 data
f3 <- fecundity %>%
  filter(adult_generation == "F3")

# plot f0 data
f0_plot <- ggplot(f0, aes(x = days_since_oviposition, y = mean)) +
  geom_line(aes(x = days_since_oviposition, y = mean, color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = T) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.8, 
                linewidth = 2, 
                show.legend = T) +
  geom_point(aes(x = days_since_oviposition, y = mean, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 5, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = T) + 
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_x_continuous(breaks = seq(from = 0, to = 28, by = 2)) +
  scale_y_continuous(breaks = seq(from = 0, to = 100, by = 20), limits = c(0,100)) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_blank()) +
  theme(legend.position = c(0.04,0.8)) +
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(-1, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(5, 10, 1, 10)) +
  guides(color = guide_legend(nrow = 1, byrow = T)) +
  labs(y = "Mean fecundity", color = "Background", fill = "Background") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f0 data
f1_plot <- ggplot(f1, aes(x = days_since_oviposition, y = mean)) +
  geom_line(aes(x = days_since_oviposition, y = mean, color = background), 
            position = position_dodge(width = 0.4),
            linewidth = 2,
            show.legend = F) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = background), 
                position = position_dodge(width = 0.4),
                width = 0.8, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = days_since_oviposition, y = mean, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 5, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = F) + 
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_x_continuous(breaks = seq(from = 0, to = 28, by = 2)) +
  scale_y_continuous(breaks = seq(from = 0, to = 100, by = 20), limits = c(0,100)) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_blank()) +
  labs(y = "Mean fecundity") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f0 data
f2_plot <- ggplot(f2, aes(x = days_since_oviposition, y = mean)) +
  geom_line(aes(x = days_since_oviposition, y = mean, color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = F) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.8, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = days_since_oviposition, y = mean, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 5, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = F) + 
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_x_continuous(breaks = seq(from = 0, to = 28, by = 2)) +
  scale_y_continuous(breaks = seq(from = 0, to = 100, by = 20), limits = c(0,100)) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_blank()) +
  labs(y = "Mean fecundity") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f0 data
f3_plot <- ggplot(f3, aes(x = days_since_oviposition, y = mean)) +
  geom_line(aes(x = days_since_oviposition, y = mean, color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = F) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.8, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = days_since_oviposition, y = mean, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 5, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = F) + 
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_x_continuous(breaks = seq(from = 0, to = 28, by = 2)) +
  scale_y_continuous(breaks = seq(from = 0, to = 100, by = 20), limits = c(0,100)) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  labs(x = "Time since first oviposition (days)", y = "Mean fecundity") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot average daily fecundity
jpeg("figures/daily_fecundity.jpeg", width = 20000, height = 30000, res = 1700)

f0_plot + 
  f1_plot + 
  f2_plot + 
  f3_plot + 
  plot_layout(nrow = 4, ncol = 1)

dev.off()
```