---
title: "Analysis 2: Lethal dose curve analysis"
author: "Joe Gunn"
date: "2024-09-19"
output: html_document
---

# <b>Project:</b> Assessing phenotypic response and transgenerational inheritance following insecticide exposure in Colorado Potato Beetles (<i>Leptinotarsa decemlineata</i>; CPB) 
<font size="+1"> We assessed inter- and trans-generational effects of sublethal insecticide exposure on four larval performance phenotypes (fecundity, herbivory, mobility, and development) in Colorado potato beetles (<i>Leptinotarsa decemlineata</i>; CPB). We implemented a replicated inbred full-sibling pedigree design, beginning with 12 "founder" mated pair lineages and continuing for four successive generations (full siblings within lineages were inbred in each generation). For the transgenerational experiment, F<sub>0</sub> larvae (offspring of founder pairs) were divided into two "background" groups (exposed, i.e., exposed to a sub-lethal [10-ppm] dose of the neonicotinoid insecticide imidacloprid; and control, i.e., exposed to an equivalent dose of molecular-grade water). All mated pair beetles across lineages, treatments, and generations were monitored for daily fecundity. A subset of third instar larvae were collected from each treatment within each lineage for each generation and monitored for three  performance traits, including larval movement (mobility), feeding rate (herbivory), and time to discrete life stages (development). Performance traits were monitored in three successive generations (F<sub>1</sub>-F<sub>3</sub>) without treatment exposure and compared to the F<sub>0</sub> generation to test for inter- and transgenerationally-inherited "memory" of phenotypic variation. An additional subset of larvae from each generation were re-exposed to the treatments, both experimental (insecticide-exposed) and control (water), to test for potential inheritance of a phenotypic "response" to insecticide stress.</font>

# <b>Specific Aim:</b> Determination of lethal concentration dose for imidacloprid insecticide treatment of the transgenerational experiment
In this analysis, we determine the sub-lethal dose concentration of insecticide (imidacloprid; see `cpb_tei_experimental_protocol.Rmd`, Figure 2) to be used as the experimental treatment (stress) in the CPB TEI experiment. We aim to use an LD25 dose of imidacloprid, which is the dosage concentration that is lethal for an average of 25% of larvae in the stressed population.

## Phases of Analysis
### Phase 1: Determination of sub-lethal (LD<sub>25</sub>) insecticide dosage

### Libraries needed for analysis
```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(writexl)
```

## PHASE 1: DETERMINATION OF SUB-LETHAL (LD<sub>25</sub>) INSECTICIDE DOSAGE
In this phase of analysis, we use 24-hour survival data from a insecticide dosage experiment (see `cpb_tei_experimental_protocol.Rmd`, Section 3) conducted on I3 larvae collected from the VTM CPB colony (see `cpb_tei_experimental_protocol.Rmd`, Section 1, for description of colony origins and laboratory maintenance). Briefly, we used a logistic regression analysis framework with a probit link function to assess the probability of survival of I3 larvae at different pre-set concentrations of insecticide (i.e., imidacloprid).

We conducted the LD assay (as described in the experimental protocol) over multiple days over the course of two months, depending on the availability of I3 larvae in the VTM laboratory colony. We therefore randomized all individual larvae over dosage treatments and over multiple 6-well assay plates. Specifically, we generated a list of individual larva IDs to reflect colony, dosage treatment received, and a unique numerical identifier,  based on the number of samples needed per dosage treatment. These IDs were then randomized onto 50 6-well plates.

### STEP 1: Read in, filter, and clean full dataset
In this step, we read in the raw data file (which includes some sample metadata, described below; `../raw_data/ld_data.xlsx`), manipulate and clean the data.

<b>Metadata</b>:
   1. "assay": original assay vs. tests of calculated LD<sub>25</sub> dosages
   2. "assay_date": date on which assay was conducted
   3. "dose_time": time of day when individuals were treated and assay was started
   4. "pipettor_name": name of researcher pipetting aqueous imidacloprid treatment
   5. "observer_name": name of researcher observing survival status
   6. "plate_number": numerical ID of 6-well plate for individual larva
   7. "well_number": numerical ID of 6-well plate for individual larva per plate
   8. "colony": alphabetic identifier of laboratory colony name (i.e., "VTM")
   9. "insecticide_date": date on which insecticide (imidacloprid) was mixed in aqueous solution
   10. "dosage": dosage (in parts-per-million, ppm) of imidacloprid (note: 0 = milliQ water)
   11. "ind_id": numerical identifier (three digits) for individual larva (001, 002, ..., 300)
   12. "molted": "yes/no" indicating whether the larva molted ("yes") or did not molt ("no") during the assay
   13. "survival_status": binary (1/0) indicator of whether individual was alive (i.e., "1") or dead (i.e., "0") after 48-hours

There are additional columns in this dataset which correspond to an independent project assessing larval mobility across different insecticide doses. These columns are removed from the dataset for ld assay analyses. 
   
#### 1a: Read in raw data, which includes all original LD assay data and additional test data, and clean for downstream analyses; run the Rmd chunk below

##### Read in and clean raw data
```{r}
# Read in raw data file (ld)
ld <- read_excel("../raw_data/ld_data.xlsx")

# Remove columns that are not needed for ld analysis (columns after "survival_status")
ld <- ld %>%
  select(assay:survival_status)

# Convert characters to factors for categorical data and drop NA (missing values)
ld <- ld %>%
  mutate(assay = factor(assay)) %>%
  mutate(pipettor_name = factor(pipettor_name)) %>%
  mutate(observer_name = factor(observer_name)) %>%
  mutate(plate_number = factor(plate_number)) %>%
  mutate(colony = factor(colony)) %>%
  mutate(insecticide_date = factor(insecticide_date)) %>%
  mutate(ind_id = factor(ind_id)) %>%
  mutate(molted = factor(molted)) %>%
  mutate(survival_status = as.numeric(survival_status)) %>% 
  drop_na()

# Save raw ld data
save(ld, file = "data/ld.Rda")
```

### STEP 2: Summarize data 
In this step, we summarize the survival data by dosage treatment by "averaging" survival by dosage (average of 1s and 0s will be a decimal between 0-1). We then graphically visualize the overall trend in survival probability over insecticide dosage treatments.

#### 2a: Count data in each dosage (treatment) group for data summarization; run the Rmd chunk below:
In this step, we organize data by dosage treatment and verify sample sizes (<i>N</i> = 30 per treatment).

##### Count data in dosage treatment groups:
```{r}
# Load in initial LD data
load("data/ld.Rda")

# Count total number of samples
ld %>%
  count()

# Count number of samples per treatment group
ld %>%
  group_by(dosage) %>%
  count()
```

<b>Filtering results</b>:
N<sub>samples</sub> = 299

A single larva was missing in the 1-ppm dosage treatment group and thus was removed from analyses (drop_na(), Step 1b above).

#### 2b: Group data by dosage and average survival probability; run the Rmd chunk below.

##### Summarize data and visualize survival over dosages: `figures/survival.pdf`
```{r}
# Load in initial LD data
load("data/ld.Rda")

# Calculate "average survival" by averaging 1s (alive) and 0s (dead) at different dosages
ld_averages <- ld %>%
  mutate(dosage = factor(dosage)) %>%
  group_by(dosage) %>%
  summarize(mean = mean(survival_status))

# Plot survival for each dosage
jpeg("figures/survival.jpeg", width = 5000, height = 4000, res = 700)

ggplot(ld_averages, aes(x = dosage, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge(), fill = "grey", color = "black", linewidth = 1) +
  geom_text(aes(label = paste(format(round(mean*100, digits = 2), nsmall =2), "%"), vjust = -0.4, hjust = 0.5, size = 4), show.legend = F) +
  theme_set(theme_cowplot(12)) +
  labs(x = "Dosage (ppm)", y = "Average 48-hr survival proportion") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(axis.title = element_text(size = 15)) +
  theme(axis.text = element_text(size = 15)) +
  theme(axis.ticks.x = element_blank()) +
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) 
  
dev.off()
```

### STEP 3: Fit logistic regression analysis with a probit link function
In this step, we fit the data using logistic regression with a probit link function to assess the probability of survival across insecticide dosage treatments.

#### 3a: Run probit analysis; run the Rmd chunk below.

##### Run probit logistic regression analysis
```{r}
# Load in raw LD data
load("data/ld.Rda")

# manipulate zeros (control) so that they can be included in og-transformed data (log10) by adding 1 to all values
ld_probit <- ld %>%
  mutate(log_dosage = log10(dosage + 1))

# Run probit analysis (logistic regression with probit link function) using glm()
probit <- glm(survival_status ~ log_dosage, 
              data = ld_probit, 
              family = binomial(link = "probit"))

# Get coefficients of probit function
summary(probit)

# Define a function to extract the ld25 value (dosage at which 75% of individuals survive on average)
findInt <- function(model, value) {
    function(x) {
        predict(model, data.frame(log_dosage = x), 
                type = "response") - value
    }
}

# Extract ld25 value
uniroot(findInt(probit, 0.75), 
        range(ld_probit$log_dosage))$root

# save ld probit data with log transformation
save(ld_probit, file = "data/ld_probit.Rda")
```

<b>Summary results</b>:
LD<sub>25</sub> (initial estimate) = 12.21458 ppm

#### 3b: Plot fitted probit regression for ld data; run the Rmd chunk below.

##### Plot probit regression: `figures/ld.pdf`
```{r}
# Load in log probit data
load("data/ld_probit.Rda")

# Plot probit curve of survival at all dosages
jpeg("figures/ld.jpeg", width = 5000, height = 4000, res = 800)

ggplot(ld_probit, aes(x = log_dosage, y = survival_status)) + 
  geom_point(position = position_jitter(height = 0.05), size = 4, color = "black", fill = "grey", pch = 21) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"), color = "black", fullrange = T) + 
  theme_set(theme_cowplot(12)) +
  scale_x_continuous(expand = c(0,0), limits = c(0,2)) +
  labs(x = "Dosage (ppm)", y = "Survival proportion") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(axis.title = element_text(size = 15)) +
  theme(axis.text = element_text(size = 15)) +
  theme(axis.ticks.x = element_blank()) +
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) 

dev.off()
```

### ------------------- END OF PHASE 1: DETERMINATION OF SUB-LETHAL (LD<sub>25</sub>) INSECTICIDE DOSAGE ------------------- ### 
