---
ttitle: "Analysis 6: Development analysis"
author: "Joe Gunn"
date: "2024-09-19"
output: html_document
---

# <b>Project:</b> Assessing phenotypic response and transgenerational inheritance following insecticide exposure in Colorado Potato Beetles (<i>Leptinotarsa decemlineata</i>; CPB) 
<font size="+1"> We assessed inter- and trans-generational effects of sublethal insecticide exposure on four larval performance phenotypes (fecundity, herbivory, mobility, and development) in Colorado potato beetles (<i>Leptinotarsa decemlineata</i>; CPB). We implemented a replicated inbred full-sibling pedigree design, beginning with 12 "founder" mated pair lineages and continuing for four successive generations (full siblings within lineages were inbred in each generation). For the transgenerational experiment, F<sub>0</sub> larvae (offspring of founder pairs) were divided into two "background" groups (exposed, i.e., exposed to a sub-lethal [10-ppm] dose of the neonicotinoid insecticide imidacloprid; and control, i.e., exposed to an equivalent dose of molecular-grade water). All mated pair beetles across lineages, treatments, and generations were monitored for daily fecundity. A subset of third instar larvae were collected from each treatment within each lineage for each generation and monitored for three  performance traits, including larval movement (mobility), feeding rate (herbivory), and time to discrete life stages (development). Performance traits were monitored in three successive generations (F<sub>1</sub>-F<sub>3</sub>) without treatment exposure and compared to the F<sub>0</sub> generation to test for inter- and transgenerationally-inherited "memory" of phenotypic variation. An additional subset of larvae from each generation were re-exposed to the treatments, both experimental (insecticide-exposed) and control (water), to test for potential inheritance of a phenotypic "response" to insecticide stress.</font>

# <b>Specific Aim:</b> Development analysis
<font size="+1">In this analysis, we assess differences in larval development (i.e., time to pupation, time from pupae to adult emergence, pupal mass, and adult mass) among mated pairs in the transgenerational experiment by generation (i.e., F0, F1, F2, F3), lineage (i.e., L1, L2, ... L10), and treatment (i.e., control and exposed). </font>

## Phases of Analysis
### Phase 1: Data summary statistics
### Phase 2: Development analysis

### Libraries needed for analysis
```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(writexl)
library(rstatix)
library(lme4)
library(rsq)
library(scales)
library(fitdistrplus)
library(forcats)
library(patchwork)
library(glmmTMB)
```

## PHASE 1: DATA SUMMARY STATISTICS
In this phase of the analysis, we calculate mean, standard deviation, and variance for all metrics (including also pupal mass and adult mass) within experimental groups of interest.

### STEP 1: Test for pairwise correlations between development metrics.
In this step, we test whether development metrics with similar units and measured in the same individuals (i.e., pupal mass vs. adult mass; time to pupation vs. pupation duration) are tightly correlated. Assessing differences between groups in highly correlated variables is redundant for addressing experimental questions. In the event that multiple variables are highly correlated, they do not all need to be analyzed.

#### 1a: Test for correlations between each pair of variables using a linear mixed effect model framework; run the Rmd chunk below.

##### Test for correlations between variables:
```{r}
# load development data
load("data/development.Rda")

# drop NAs and recalculate metrics for development time (time to pupation and pupa duration)
development <- development %>%
  drop_na(pupa_date:adult_mass) %>%
  mutate(pupa_time = as.integer(difftime(pupa_date, assay_date, "days")),
         pupa_duration = as.integer(difftime(emerge_date, pupa_date, "days")))

# test for correlation between total distance walked and average speed
mass <- lmer(pupa_mass ~ adult_mass + (1|lineage),
             data = development)

# test for correlation between total distance walked and average acceleration
time <- lmer(pupa_time ~ pupa_duration + (1|lineage),
             data = development)

# summarize model results for mass metrics
anova(mass) # get F values
Anova(mass) # get p-values
rsq.lmm(mass) # get r-squared values

# summarize model results for time metrics
anova(time) # get F values
Anova(time) # get p-values
rsq.lmm(time) # get r-squared values
```

<b>Analysis results:</b> <br>

<b>Coefficients of linear mixed effect models for pairwise metric comparisons:</b><br>
<b>pupa mass and adult mass: F = 696.17; p < 0.0001; r-squared = 0.7527678 </b><br>
<b>pupation time and pupa duration: F = 3.0383; p = 0.08132; r-squared = 0.06688807 </b><br>

Given that pupal and adult mass are are significantly correlated, we opted to use only a single metric as a proxy for individual mass, i.e., pupa mass (we chose pupa mass since many individuals that pupated did not make it to adult emergence and we therefore had a larger dataset). Time to pupation and duration of pupation were not significantly correlated, so we analyzed both of these metrics. Downstream analyses were therefore conducted for pupa mass, pupa time, and pupa duration.

#### 1b: Plot correlations; run the Rmd chunk below.

##### Plot correlations: `figures/correlations.jpeg`
```{r}
# load development data
load("data/development.Rda")

# drop NAs and recalculate metrics for development time (time to pupation and pupa duration)
development <- development %>%
  drop_na(pupa_date:adult_mass) %>%
  mutate(pupa_time = as.integer(difftime(pupa_date, assay_date, "days")),
         pupa_duration = as.integer(difftime(emerge_date, pupa_date, "days")))

# plot mass correlation
mass <- ggplot(development, aes(x = pupa_mass, y = adult_mass)) + 
  geom_point(fill = "grey", show.legend = F, size = 3, pch = 21, color = "black") + 
  geom_smooth(method = "lm", color = "black", fullrange = T) +
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("grey")) +
  scale_color_manual(values = c("black")) + 
  xlim(0,200) +
  ylim(0,200) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  scale_x_continuous(labels = number_format(accuracy = 0.1), expand = c(0,0)) + 
  scale_y_continuous(labels = number_format(accuracy = 0.1), expand = c(0,0)) +
  labs(x = "Pupa mass (mg)", y = "Adult mass (mg)") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot time correlation
time <- ggplot(development, aes(x = pupa_time, y = pupa_duration)) + 
  geom_point(fill = "grey", show.legend = F, size = 3, pch = 21, color = "black", position = position_jitter()) + 
  geom_smooth(method = "lm", color = "black", fullrange = T) +
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("grey")) +
  scale_color_manual(values = c("black")) + 
  xlim(0,35) +
  ylim(0,12) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  scale_x_continuous(expand = c(0,0)) + 
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "Time to pupation (days)", y = "Pupa duration (days)") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot all together
jpeg("figures/correlations.jpeg", width = 18000, height = 8000, res = 1000)

mass + time + plot_layout(nrow = 1, ncol = 2)

dev.off()
```

### STEP 2: Calculate summary statistics for larval development data; run the Rmd chunk below.
In this step, we calculate mean, standard deviation, and variance for larval development metrics (i.e., time to pupation, pupation duration, pupal mass, and adult mass) for all experimental groups.

##### Calculate summary statistics for larval development data:
```{r}
# load pupa time data
load("data/pupa_time.Rda")

# load pupa duration data
load("data/pupa_duration.Rda")

# load pupa mass data
load("data/pupa_mass.Rda")

# get sample size (n), mean, standard deviation (sd), and standard error (se) for time to pupation by generation and treatment
pupa_time %>%
  group_by(larval_generation, background, treatment) %>% 
  summarize(n = n(),
            mean = mean(pupa_time),
            sd = sd(pupa_time),
            se = sd/sqrt(n()))

# get sample size (n), mean, standard deviation (sd), and standard error (se) for pupation duration by generation and treatment
pupa_duration %>%
  group_by(larval_generation, background, treatment) %>% 
  summarize(n = n(),
            mean = mean(pupa_duration),
            sd = sd(pupa_duration),
            se = sd/sqrt(n()))

# get sample size (n), mean, standard deviation (sd), and standard error (se) for pupation duration by generation and treatment
pupa_mass %>%
  group_by(larval_generation, background, treatment) %>% 
  summarize(n = n(),
            mean = mean(pupa_mass),
            sd = sd(pupa_mass),
            se = sd/sqrt(n()))
```

These data are presented in Table 1 of the final ms.
