---
title: "extra"
output: html_document
---

### STEP 2: Assess phenotypic memory and response in average walking speed.
In this step, we assess phenotypic memory and response in average walking speed by larvae.

#### 2a: Test fit of empirical data to a normal distribution; run the Rmd chunk below.
In this step, we use the shapiro-wilk test to assess whether average walking speed among larvae in each group of interest deviates from a normal distribution.

##### Test fit of normality:
```{r}
# load mobility data
load("data/mobility.Rda")

# set to show decimals rather than scientific notation for p-value assessments
options(scipen = 999)

# test for fit of normality with treatments and generations and compare to alpha of 0.05/14 = 0.003 (to account for multiple comparisons)
mobility %>%
  group_by(larval_generation, background, treatment) %>%
  shapiro_test(ave_speed) %>%
  mutate(alpha = 0.003) %>%
  mutate(deviates = ifelse(p < alpha, "yes", "no"))
```

<b>Analysis results:</b> <br>

<b>Shapiro-wilk p-values for each treatment and generation:</b><br>
<b>F0_C_C: p < 0.001 </b><br>
<b>F0_C_E: p < 0.001 </b><br>
<b>F1_C_C: p < 0.001 </b><br>
<b>F1_C_E: p = 0.010 </b><br>
<b>F1_E_C: p = 0.002 </b><br>
<b>F1_E_E: p < 0.001 </b><br>
<b>F2_C_C: p = 0.180 </b><br>
<b>F2_C_E: p = 0.010 </b><br>
<b>F2_E_C: p = 0.082 </b><br>
<b>F2_E_E: p = 0.199 </b><br>
<b>F3_C_C: p < 0.001 </b><br>
<b>F3_C_E: p = 0.012 </b><br>
<b>F3_E_C: p = 0.011 </b><br>
<b>F3_E_E: p < 0.001 </b><br>

Since 7/14 tests for normality fit indicate a deviation from a normal distribution, we apply a square-root transformation to the raw data to prepare for analysis.

#### 2b: Square-root transform average walking speed data and re-assess fit of normal distribution; run the Rmd chunk below.
In this step, we calculate the square root of the raw average speed data and re-assess the fit of the transformed data to a normal distribution.

##### Transform data and re-assess fit of normality:
```{r}
# load mobility data
load("data/mobility.Rda")

# set to show decimals rather than scientific notation for p-value assessments
options(scipen = 999)

# square root transform walking speed data and add to dataset
speed <- mobility %>%
  mutate(sqrt_speed = sqrt(ave_speed))

# re-assess fit of normality to square-root-transformed data (based on p-value correction for multiple tests)
speed %>%
  group_by(larval_generation, background, treatment) %>%
  shapiro_test(sqrt_speed) %>%
  mutate(alpha = 0.003) %>%
  mutate(deviates = ifelse(p < alpha, "yes", "no"))

# save transformed speed mobility data
save(speed, file = "data/speed.Rda")
```

<b>Analysis results:</b> <br>

<b>Shapiro-wilk p-values for each treatment and generation:</b><br>
<b>F0_C_C: p = 0.436 </b><br>
<b>F0_C_E: p = 0.008 </b><br>
<b>F1_C_C: p = 0.289 </b><br>
<b>F1_C_E: p = 0.394 </b><br>
<b>F1_E_C: p = 0.354 </b><br>
<b>F1_E_E: p = 0.103 </b><br>
<b>F2_C_C: p = 0.423 </b><br>
<b>F2_C_E: p = 0.343 </b><br>
<b>F2_E_C: p = 0.577 </b><br>
<b>F2_E_E: p = 0.094 </b><br>
<b>F3_C_C: p = 0.100 </b><br>
<b>F3_C_E: p = 0.548 </b><br>
<b>F3_E_C: p = 0.169 </b><br>
<b>F3_E_E: p = 0.159 </b><br>

All 14 tests indicate fit of the normal distribution, so we continue with analysis using these transformed data.

#### 2c: Assess phenotypic memory of insecticide stress in average walking speed; run the Rmd chunk below.

##### Assess phenotypic memory of insecticide stress in average walking speed:
```{r}

```

#### 2d: Plot transformed data for average walking speed to visualize phenotypic memory; run the Rmd chunk below.

##### Plot transformed data for phenotypic memory analysis: `figures/speed_memory.jpeg`
```{r}
# load speed mobility data
load("data/speed.Rda")

# get only individuals given the control in each generation (for both background exposure groups)
speed <- speed %>%
  filter(treatment == "C")

# calculate means, standard deviation, and standard error of square-root transformed data 
speed <- speed %>%
  group_by(larval_generation, background) %>%
  summarize(mean = mean(sqrt_speed),
            sd = sd(sqrt_speed),
            n = n(),
            se = sd/sqrt(n))

# plot transformed data by background, generation, and treatment
jpeg("figures/speed_memory.jpeg", width = 10000, height = 8000, res = 700)
 
ggplot(speed, aes(x = larval_generation, y = mean, fill = background)) + 
  geom_bar(position = position_dodge(), stat = "identity", show.legend = T, color = "black", linewidth = 1.5) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.4, position = position_dodge(0.9), linewidth = 1.5) + 
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 0.8, by = 0.2)) +
  ylim(0,0.8) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 40, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) +
  theme(legend.position = c(0.05,0.87)) + 
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(0.2, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(10, 10, 10, 10)) +
  labs(x = "Larval generation", y = "Square root of walking speed (mm/s)", fill = "Background") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(15,15,15,15))

dev.off()
```

#### 2e: Assess phenotypic response to insecticide stress in average walking speed; run the Rmd chunk below.

##### Assess phenotypic response to insecticide stress in average walking speed:
```{r}

```

#### 2f: Plot transformed data for average walking speed to visualize phenotypic response; run the Rmd chunk below.

##### Plot transformed data for phenotypic response analysis (for both control and exposed background samples): 1. `figures/speed_response_control.jpeg`; 2.`figures/speed_response_exposed.jpeg`
```{r}
# load speed mobility data
load("data/speed.Rda")

# calculate means, standard deviation, and standard error of square-root transformed data 
speed <- speed %>%
  group_by(larval_generation, background, treatment) %>%
  summarize(mean = mean(sqrt_speed),
            sd = sd(sqrt_speed),
            n = n(),
            se = sd/sqrt(n))

# Convert background to "C" for both F0 groups and change treatment to C and E for each group, respectively
speed_c <- speed
speed_c$background[2] <- "C" 
speed_c$treatment[2] <- "E"

# get only individuals with the control background
speed_c <- speed_c %>%
  filter(background == "C")

# Convert background to "E" for both F0 groups and change treatment to C and E for each group, respectively
speed_e <- speed
speed_e$background[1] <- "E" 
speed_e$treatment[2] <- "E"

# get only individuals with the exposed background
speed_e <- speed_e %>%
  filter(background == "E")

# plot transformed data by background, generation, and treatment for control background
jpeg("figures/speed_response_control.jpeg", width = 10000, height = 8000, res = 700)
 
ggplot(speed_c, aes(x = larval_generation, y = mean, fill = treatment)) + 
  geom_bar(position = position_dodge(), stat = "identity", show.legend = T, color = "black", linewidth = 1.5) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.4, position = position_dodge(0.9), linewidth = 1.5) + 
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 0.8, by = 0.2)) +
  ylim(0,0.8) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 40, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) +
  theme(legend.position = c(0.05,0.87)) + 
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(0.2, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(10, 10, 10, 10)) +
  labs(x = "Larval generation", y = "Square root of walking speed (mm/s)", fill = "Treatment") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(15,15,15,15))

dev.off()

# plot transformed data by background, generation, and treatment for exposed background
jpeg("figures/speed_response_exposed.jpeg", width = 10000, height = 8000, res = 700)
 
ggplot(speed_e, aes(x = larval_generation, y = mean, fill = treatment)) + 
  geom_bar(position = position_dodge(), stat = "identity", show.legend = T, color = "black", linewidth = 1.5) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.4, position = position_dodge(0.9), linewidth = 1.5) + 
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 0.8, by = 0.2)) +
  ylim(0,0.8) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 40, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) +
  theme(legend.position = c(0.05,0.87)) + 
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(0.2, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(10, 10, 10, 10)) +
  labs(x = "Larval generation", y = "Square root of walking speed (mm)", fill = "Treatment") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(15,15,15,15))

dev.off()
```

### STEP 3: Assess phenotypic memory and response in average walking acceleration.
In this step, we assess phenotypic memory and response in average walking acceleration by larvae.

#### 3a: Test fit of empirical data to a normal distribution; run the Rmd chunk below.
In this step, we use the shapiro-wilk test to assess whether average walking acceleration among larvae in each group of interest deviates from a normal distribution.

##### Test fit of normality:
```{r}
# load mobility data
load("data/mobility.Rda")

# set to show decimals rather than scientific notation for p-value assessments
options(scipen = 999)

# test for fit of normality with treatments and generations and compare to alpha of 0.05/14 = 0.003 (to account for multiple comparisons)
mobility %>%
  group_by(larval_generation, background, treatment) %>%
  shapiro_test(ave_accel) %>%
  mutate(alpha = 0.003) %>%
  mutate(deviates = ifelse(p < alpha, "yes", "no"))
```

<b>Analysis results:</b> <br>

<b>Shapiro-wilk p-values for each treatment and generation:</b><br>
<b>F0_C_C: p < 0.001 </b><br>
<b>F0_C_E: p < 0.001 </b><br>
<b>F1_C_C: p = 0.072 </b><br>
<b>F1_C_E: p < 0.001 </b><br>
<b>F1_E_C: p < 0.001 </b><br>
<b>F1_E_E: p < 0.001 </b><br>
<b>F2_C_C: p = 0.091 </b><br>
<b>F2_C_E: p = 0.107 </b><br>
<b>F2_E_C: p = 0.055 </b><br>
<b>F2_E_E: p = 0.891 </b><br>
<b>F3_C_C: p < 0.001 </b><br>
<b>F3_C_E: p = 0.008 </b><br>
<b>F3_E_C: p = 0.002 </b><br>
<b>F3_E_E: p = 0.001 </b><br>

Since 8/14 tests for normality fit indicate a deviation from a normal distribution, we apply a square-root transformation to the raw data to prepare for analysis.

#### 3b: Square-root transform average walking acceleration data and re-assess fit of normal distribution; run the Rmd chunk below.
In this step, we calculate the square root of the raw average acceleration data and re-assess the fit of the transformed data to a normal distribution.

##### Transform data and re-assess fit of normality:
```{r}
# load mobility data
load("data/mobility.Rda")

# set to show decimals rather than scientific notation for p-value assessments
options(scipen = 999)

# square root transform walking acceleration data and add to dataset
acceleration <- mobility %>%
  mutate(sqrt_accel = sqrt(ave_accel))

# re-assess fit of normality to square-root-transformed data (based on p-value correction for multiple tests)
acceleration %>%
  group_by(larval_generation, background, treatment) %>%
  shapiro_test(sqrt_accel) %>%
  mutate(alpha = 0.003) %>%
  mutate(deviates = ifelse(p < alpha, "yes", "no"))

# save transformed acceleration mobility data
save(acceleration, file = "data/acceleration.Rda")
```

<b>Analysis results:</b> <br>

<b>Shapiro-wilk p-values for each treatment and generation:</b><br>
<b>F0_C_C: p = 0.011 </b><br>
<b>F0_C_E: p = 0.007 </b><br>
<b>F1_C_C: p = 0.652 </b><br>
<b>F1_C_E: p = 0.595 </b><br>
<b>F1_E_C: p = 0.129 </b><br>
<b>F1_E_E: p = 0.004 </b><br>
<b>F2_C_C: p = 0.609 </b><br>
<b>F2_C_E: p = 0.157 </b><br>
<b>F2_E_C: p = 0.368 </b><br>
<b>F2_E_E: p = 0.353 </b><br>
<b>F3_C_C: p = 0.004 </b><br>
<b>F3_C_E: p = 0.296 </b><br>
<b>F3_E_C: p = 0.076 </b><br>
<b>F3_E_E: p = 0.145 </b><br>

All 14 tests indicate fit of the normal distribution, so we continue with analysis using these transformed data.

#### 3c: Assess phenotypic memory of insecticide stress in average acceleration; run the Rmd chunk below.

##### Assess phenotypic memory of insecticide stress in average acceleration:
```{r}

```

#### 3d: Plot transformed data for average acceleration to visualize phenotypic memory; run the Rmd chunk below.

##### Plot transformed data for phenotypic memory analysis: `figures/speed_memory.jpeg`
```{r}
# load speed mobility data
load("data/acceleration.Rda")

# get only individuals given the control in each generation (for both background exposure groups)
acceleration <- acceleration %>%
  filter(treatment == "C")

# calculate means, standard deviation, and standard error of square-root transformed data 
acceleration <- acceleration %>%
  group_by(larval_generation, background) %>%
  summarize(mean = mean(sqrt_accel),
            sd = sd(sqrt_accel),
            n = n(),
            se = sd/sqrt(n))

# plot transformed data by background, generation, and treatment
jpeg("figures/acceleration_memory.jpeg", width = 10000, height = 8000, res = 700)
 
ggplot(acceleration, aes(x = larval_generation, y = mean, fill = background)) + 
  geom_bar(position = position_dodge(), stat = "identity", show.legend = T, color = "black", linewidth = 1.5) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.4, position = position_dodge(0.9), linewidth = 1.5) + 
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 1.5, by = 0.5)) +
  ylim(0,1.5) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 40, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) +
  theme(legend.position = c(0.05,0.87)) + 
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(0.2, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(10, 10, 10, 10)) +
  labs(x = "Larval generation", y = "Square root of acceleration (mm/s/s)", fill = "Background") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(15,15,15,15))

dev.off()
```

#### 3e: Assess phenotypic response to insecticide stress in average acceleration; run the Rmd chunk below.

##### Assess phenotypic response to insecticide stress in average acceleration:
```{r}

```

#### 3f: Plot transformed data for average acceleration to visualize phenotypic response; run the Rmd chunk below.

##### Plot transformed data for phenotypic response analysis (for both control and exposed background samples): 1. `figures/acceleration_response_control.jpeg`; 2.`figures/acceleration_response_exposed.jpeg`
```{r}
# load acceleration mobility data
load("data/acceleration.Rda")

# calculate means, standard deviation, and standard error of square-root transformed data 
acceleration <- acceleration %>%
  group_by(larval_generation, background, treatment) %>%
  summarize(mean = mean(sqrt_accel),
            sd = sd(sqrt_accel),
            n = n(),
            se = sd/sqrt(n))

# Convert background to "C" for both F0 groups and change treatment to C and E for each group, respectively
acceleration_c <- acceleration
acceleration_c$background[2] <- "C" 
acceleration_c$treatment[2] <- "E"

# get only individuals with the control background
acceleration_c <- acceleration_c %>%
  filter(background == "C")

# Convert background to "E" for both F0 groups and change treatment to C and E for each group, respectively
acceleration_e <- acceleration
acceleration_e$background[1] <- "E" 
acceleration_e$treatment[2] <- "E"

# get only individuals with the exposed background
acceleration_e <- acceleration_e %>%
  filter(background == "E")

# plot transformed data by background, generation, and treatment for control background
jpeg("figures/acceleration_response_control.jpeg", width = 10000, height = 8000, res = 700)
 
ggplot(acceleration_c, aes(x = larval_generation, y = mean, fill = treatment)) + 
  geom_bar(position = position_dodge(), stat = "identity", show.legend = T, color = "black", linewidth = 1.5) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.4, position = position_dodge(0.9), linewidth = 1.5) + 
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 1.5, by = 0.5)) +
  ylim(0,1.5) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 40, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) +
  theme(legend.position = c(0.05,0.87)) + 
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(0.2, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(10, 10, 10, 10)) +
  labs(x = "Larval generation", y = "Square root of acceleration (mm/s/s)", fill = "Treatment") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(15,15,15,15))

dev.off()

# plot transformed data by background, generation, and treatment for exposed background
jpeg("figures/acceleration_response_exposed.jpeg", width = 10000, height = 8000, res = 700)
 
ggplot(acceleration_e, aes(x = larval_generation, y = mean, fill = treatment)) + 
  geom_bar(position = position_dodge(), stat = "identity", show.legend = T, color = "black", linewidth = 1.5) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.4, position = position_dodge(0.9), linewidth = 1.5) + 
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 1.5, by = 0.5)) +
  ylim(0,1.5) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 40, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) +
  theme(legend.position = c(0.05,0.87)) + 
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(0.2, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(10, 10, 10, 10)) +
  labs(x = "Larval generation", y = "Square root of acceleration (mm/s/s)", fill = "Treatment") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(15,15,15,15))

dev.off()
```


### Distance memory boxplot:
```{r}
jpeg("figures/distance_memory.jpeg", width = 10000, height = 8000, res = 700)
 
ggplot(distance, aes(x = larval_generation, y = mean, fill = background)) + 
  geom_bar(position = position_dodge(), stat = "identity", show.legend = T, color = "black", linewidth = 1.5) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.4, position = position_dodge(0.9), linewidth = 1.5) + 
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 15, by = 3)) +
  ylim(0,15) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 40, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) +
  theme(legend.position = c(0.05,0.87)) + 
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(0.2, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(10, 10, 10, 10)) +
  labs(x = "Larval generation", y = "Square root of distance (mm)", fill = "Background") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(15,15,15,15))

dev.off()
```

```{r}
# plot transformed data by background, generation, and treatment for control background
jpeg("figures/distance_response_control.jpeg", width = 10000, height = 8000, res = 700)
 
ggplot(distance_c, aes(x = larval_generation, y = mean, fill = treatment)) + 
  geom_bar(position = position_dodge(), stat = "identity", show.legend = T, color = "black", linewidth = 1.5) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.4, position = position_dodge(0.9), linewidth = 1.5) + 
  theme_set(theme_cowplot(12)) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 15, by = 3)) +
  ylim(0,15) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 40, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.75)) +
  theme(legend.position = c(0.05,0.87)) + 
  theme(legend.text = element_text(size = 30)) +
  theme(legend.title = element_text(size = 30, face = "bold")) +
  theme(legend.key.size = unit(1, 'cm')) +
  theme(legend.key.spacing.y = unit(0.2, 'cm')) +
  theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  theme(legend.box.margin = margin(10, 10, 10, 10)) +
  labs(x = "Larval generation", y = "Square root of distance (mm)", fill = "Treatment") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(15,15,15,15))

dev.off()
```