---
ttitle: "Analysis 4: Mobility analysis"
author: "Joe Gunn"
date: "2024-09-19"
output: html_document
---

# <b>Project:</b> Assessing phenotypic response and transgenerational inheritance following insecticide exposure in Colorado Potato Beetles (<i>Leptinotarsa decemlineata</i>; CPB) 
<font size="+1"> We assessed inter- and trans-generational effects of sublethal insecticide exposure on four larval performance phenotypes (fecundity, herbivory, mobility, and development) in Colorado potato beetles (<i>Leptinotarsa decemlineata</i>; CPB). We implemented a replicated inbred full-sibling pedigree design, beginning with 12 "founder" mated pair lineages and continuing for four successive generations (full siblings within lineages were inbred in each generation). For the transgenerational experiment, F<sub>0</sub> larvae (offspring of founder pairs) were divided into two treatment groups (exposed, i.e., exposed to a sub-lethal [10-ppm] dose of the neonicotinoid insecticide imidacloprid; and control, i.e., exposed to an equivalent dose of molecular-grade water). A subset of beetles were collected from each treatment within each lineage for each generation and monitored for four performance traits, including larval movement (mobility), feeding rate (herbivory), and time to discrete life stages (development). Performance traits were monitored in three successive generations (F<sub>1</sub>-F<sub>3</sub>) without treatment exposure and compared to the F<sub>0</sub> generation to test for inter- and transgenerationally-inherited phenotypic variation. An additional subset of larvae from each generation were re-exposed to treatments, both experimental (insecticide-exposed) and control (water), to test for potential selection of phenotypic variation across generations.</font>

# <b>Specific Aim:</b> Mobility analysis
<font size="+1">In this analysis, we assess differences in larval mobility (i.e., total distance walked, average walking rate, and average walking acceleration) among mated pairs in the transgenerational experiment by generation (i.e., F0, F1, F2, F3), lineage (i.e., L1, L2, ... L10), and treatment (i.e., control and exposed). </font>

## Phases of Analysis
### Phase 1: Data Summary Statistics
### Phase 2: 

### Libraries needed for analysis:
```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(writexl)
```

## PHASE 1: DATA SUMMARY STATISTICS
In this phase of the analysis, we calculate mean, standard deviation, and variance for mobility metrics within experimental groups of interest.

### STEP 1: Calculate summary statistics for fecundity data; run the Rmd chunk below.
In this step, we calculate mean, standard deviation, and variance for mobility (i.e., walking speed, walking acceleration, and total distance walked) for all experimental groups.

##### Calculate summary statistics for fecundity data:
```{r}
# load mobility data
load("data/mobility.Rda")

# get sample size (n), mean, standard deviation (sd), and standard error (se) for walking rate by generation and treatment
mobility %>%
  group_by(larval_generation, background, treatment) %>% 
  summarize(n = n(),
            mean = mean(ave_speed),
            sd = sd(ave_speed),
            se = sd/sqrt(n()))

# get sample size (n), mean, standard deviation (sd), and standard error (se) for average acceleration by generation and treatment
mobility %>%
  group_by(larval_generation, background, treatment) %>% 
  summarize(n = n(),
            mean = mean(ave_accel),
            sd = sd(ave_accel),
            se = sd/sqrt(n()))

# get sample size (n), mean, standard deviation (sd), and standard error (se) for distance walked by generation and treatment
mobility %>%
  group_by(larval_generation, background, treatment) %>% 
  summarize(n = n(),
            mean = mean(total_distance),
            sd = sd(total_distance),
            se = sd/sqrt(n()))
```
