---
ttitle: "Analysis 5: Herbivory analysis"
author: "Joe Gunn"
date: "2024-09-19"
output: html_document
---

# <b>Project:</b> Assessing phenotypic response and transgenerational inheritance following insecticide exposure in Colorado Potato Beetles (<i>Leptinotarsa decemlineata</i>; CPB) 
<font size="+1"> We assessed inter- and trans-generational effects of sublethal insecticide exposure on four larval performance phenotypes (fecundity, herbivory, mobility, and development) in Colorado potato beetles (<i>Leptinotarsa decemlineata</i>; CPB). We implemented a replicated inbred full-sibling pedigree design, beginning with 12 "founder" mated pair lineages and continuing for four successive generations (full siblings within lineages were inbred in each generation). For the transgenerational experiment, F<sub>0</sub> larvae (offspring of founder pairs) were divided into two "background" groups (exposed, i.e., exposed to a sub-lethal [10-ppm] dose of the neonicotinoid insecticide imidacloprid; and control, i.e., exposed to an equivalent dose of molecular-grade water). All mated pair beetles across lineages, treatments, and generations were monitored for daily fecundity. A subset of third instar larvae were collected from each treatment within each lineage for each generation and monitored for three  performance traits, including larval movement (mobility), feeding rate (herbivory), and time to discrete life stages (development). Performance traits were monitored in three successive generations (F<sub>1</sub>-F<sub>3</sub>) without treatment exposure and compared to the F<sub>0</sub> generation to test for inter- and transgenerationally-inherited "memory" of phenotypic variation. An additional subset of larvae from each generation were re-exposed to the treatments, both experimental (insecticide-exposed) and control (water), to test for potential inheritance of a phenotypic "response" to insecticide stress.</font>

# <b>Specific Aim:</b> Herbivory analysis
<font size="+1">In this analysis, we assess differences in larval herbivory (i.e., total leaf tissue consumed) among mated pairs in the transgenerational experiment by generation (i.e., F0, F1, F2, F3), lineage (i.e., L1, L2, ... L10), and treatment (i.e., control and exposed). </font>

## Phases of Analysis
### Phase 1: Data Summary Statistics
### Phase 2: 

### Libraries needed for analysis:
```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(writexl)
library(rstatix)
library(lme4)
library(rsq)
library(scales)
library(lmerTest)
library(fitdistrplus)
library(forcats)
```

## PHASE 1: DATA SUMMARY STATISTICS
In this phase of the analysis, we calculate mean, standard deviation, and variance for larval herbivory metrics within experimental groups of interest.

### STEP 1: Calculate summary statistics for larval herbivory data; run the Rmd chunk below.
In this step, we calculate mean, standard deviation, and variance for larval herbivory (i.e., total leaf area consumed) for all experimental groups.

##### Calculate summary statistics for larval herbivory data:
```{r}
# load herbivory data
load("data/herbivory.Rda")

herbivory <- herbivory %>%
  mutate(consumed = as.numeric(ifelse(leaf_area_consumed != 0, "1", "0"))) %>%
  group_by(larval_generation, background, treatment) %>%
  summarize(sum_consumed = sum(consumed),
            n = n(),
            prop_consumed = sum_consumed/n,
            mean_consumed = mean(leaf_area_consumed),
            sd_consumed = sd(leaf_area_consumed),
            se_consumed = sd_consumed/sqrt(n))

# get sample size (n), mean, standard deviation (sd), and standard error (se) for walking rate by generation and treatment
herbivory %>%
  group_by(larval_generation, background, treatment) %>% 
  summarize(n = n(),
            mean = mean(total_leaf_area),
            sd = sd(total_leaf_area),
            se = sd/sqrt(n()))


f1 <- herbivory %>%
  filter(larval_generation == "F1")

model = glmer(consumed ~ background*treatment + (1|lineage), 
            family = 'binomial',
            data = f1)

summary(model)


dat = as.data.frame(Titanic)
dat = dat[rep(1:nrow(dat), dat$Freq), -5]
```
## PHASE 2: HERBIVORY ANALYSIS
In this phase of the analysis, we assess differences in larval herbivory (i.e., total leaf area consumed) between treatments and background exposure statuses across generations, accounting for variation in lineages as a random effect in a linear mixed-effect model framework. We assess two distinct modes of potential phenotypic effects: 1) assessing inheritance of phenotypic "memory" of stress by comparing phenotypic variation (i.e., herbivory metrics) in beetles with a background (F0 exposure) of insecticide vs. those with a background of the water control and without subsequent exposure in successive generations; and 2) assessing inheritance of a phenotypic "response" to stress by comparing phenotypic variation (i.e., herbivory metrics) in beetles treated in each generation with insecticide or a water control and which had a background (F0 exposure) of either insecticide or the water control. 

### STEP 1: Test fit of empirical data to a normal distribution; run the Rmd chunk below.
In this step, we use the shapiro-wilk test to assess whether total leaf area consumed among larvae in each group of interest deviates from a normal distribution.

##### Test fit of normality:
```{r}
# load herbivory data
load("data/herbivory.Rda")

# set to show decimals rather than scientific notation for p-value assessments
options(scipen = 999)

# test for fit of normality with treatments and generations and compare to alpha of 0.05/14 = 0.003 (to account for multiple comparisons)
herbivory %>%
  group_by(larval_generation, background, treatment) %>%
  shapiro_test(leaf_area_consumed) %>%
  mutate(alpha = 0.003) %>%
  mutate(deviates = ifelse(p < alpha, "yes", "no"))
```

<b>Analysis results:</b> <br>

<b>Shapiro-wilk p-values for each treatment and generation:</b><br>
<b>F0_C_N: p < 0.001 </b><br>
<b>F0_E_N: p < 0.001 </b><br>
<b>F1_C_N: p = 0.015 </b><br>
<b>F1_C_R: p < 0.001</b><br>
<b>F1_E_N: p = 0.005 </b><br>
<b>F1_E_R: p < 0.001 </b><br>
<b>F2_C_N: p = 0.209 </b><br>
<b>F2_C_R: p < 0.001 </b><br>
<b>F2_E_N: p = 0.007 </b><br>
<b>F2_E_R: p < 0.001</b><br>
<b>F3_C_N: p < 0.001 </b><br>
<b>F3_C_R: p < 0.001 </b><br>
<b>F3_E_N: p < 0.001 </b><br>
<b>F3_E_R: p < 0.001 </b><br>

Since 10/14 tests indicate a deviation from a normal distribution, we test the fit of multiple alternative distributions to determine whether there is consistent best-fitting distribution for the data across groups of interest.

### STEP 2: Test fit of empirical data to alternative distributions; run the Rmd chunk below.
In this step, we test the fit of three additional distributions (in addition to the normal distribution for comparison), including uniform, negative binomial, and poisson distributions, and determine the best fitting distribution for each group of interest. 

##### Test fit of alternative distributions:
```{r}
# load herbivory data
load("data/herbivory.Rda")

# set to show decimals rather than scientific notation for p-value assessments
options(scipen = 999)

# Generate combined "group" factor, which includes larval generation, background, and treatment, to test distributions within each group of interest.
herbivory <- herbivory %>%
  unite(group, larval_generation, c(background, treatment), sep = "_", remove = F) %>%
  mutate(group = factor(group))

# generate list of levels within group factor
levels <- levels(herbivory$group)

# define empty lists to hold statistics for each distribution test
unif <- list() 
norm <- list()
nbin <- list() 
pois <- list()

# Write for loop to fit each distribution to data, iterating through group levels (i)
for(i in levels) {
  
  # generate subsets of data for each group level
  subset <- herbivory[herbivory$group == i, ]
  
  # fit distributions to leaf area for each subset
  fit_unif <- fitdist(subset$leaf_area_consumed, "unif") 
  fit_norm <- fitdist(subset$leaf_area_consumed, "norm") 
  fit_nbin <- fitdist(subset$leaf_area_consumed, "nbinom") 
  fit_pois <- fitdist(subset$leaf_area_consumed, "pois")
  
  # fill list with fit results for each distribution
  unif[[i]] <- fit_unif 
  norm[[i]] <- fit_norm 
  nbin[[i]] <- fit_nbin 
  pois[[i]] <- fit_pois
  
}

# Get table of distribution stats for F0 C N group
stats_1 <- data_frame(F0_C_N = gofstat(list(unif$F0_C_N, 
                                            norm$F0_C_N, 
                                            nbin$F0_C_N,
                                            pois$F0_C_N))$aic)

# Get table of distribution stats for F0 E N group
stats_2 <- data_frame(F0_E_N = gofstat(list(unif$F0_E_N, 
                                            norm$F0_E_N, 
                                            nbin$F0_E_N,
                                            pois$F0_E_N))$aic)

# Get table of distribution stats for F1 C N group
stats_3 <- data_frame(F1_C_N = gofstat(list(unif$F1_C_N, 
                                            norm$F1_C_N, 
                                            nbin$F1_C_N,
                                            pois$F1_C_N))$aic)

# Get table of distribution stats for F1 C R group
stats_4 <- data_frame(F1_C_R = gofstat(list(unif$F1_C_R, 
                                            norm$F1_C_R, 
                                            nbin$F1_C_R,
                                            pois$F1_C_R))$aic)
         
# Get table of distribution stats for F1 E N group
stats_5 <- data_frame(F1_E_N = gofstat(list(unif$F1_E_N, 
                                            norm$F1_E_N, 
                                            nbin$F1_E_N,
                                            pois$F1_E_N))$aic)

# Get table of distribution stats for F1 E R group
stats_6 <- data_frame(F1_E_R = gofstat(list(unif$F1_E_R, 
                                            norm$F1_E_R, 
                                            nbin$F1_E_R,
                                            pois$F1_E_R))$aic)

# Get table of distribution stats for F2 C N group
stats_7 <- data_frame(F2_C_N = gofstat(list(unif$F2_C_N, 
                                            norm$F2_C_N, 
                                            nbin$F2_C_N,
                                            pois$F2_C_N))$aic)

# Get table of distribution stats for F2 C R group
stats_8 <- data_frame(F2_C_R = gofstat(list(unif$F2_C_R, 
                                            norm$F2_C_R, 
                                            nbin$F2_C_R,
                                            pois$F2_C_R))$aic)

# Get table of distribution stats for F2 E N group
stats_9 <- data_frame(F2_E_N = gofstat(list(unif$F2_E_N, 
                                            norm$F2_E_N, 
                                            nbin$F2_E_N,
                                            pois$F2_E_N))$aic)

# Get table of distribution stats for F2 E R group
stats_10 <- data_frame(F2_E_R = gofstat(list(unif$F2_E_R, 
                                             norm$F2_E_R, 
                                             nbin$F2_E_R,
                                             pois$F2_E_R))$aic)

# Get table of distribution stats for F3 C N group
stats_11 <- data_frame(F3_C_N = gofstat(list(unif$F3_C_N, 
                                             norm$F3_C_N, 
                                             nbin$F3_C_N,
                                             pois$F3_C_N))$aic)

# Get table of distribution stats for F3 C R group
stats_12 <- data_frame(F3_C_R = gofstat(list(unif$F3_C_R, 
                                             norm$F3_C_R, 
                                             nbin$F3_C_R,
                                             pois$F3_C_R))$aic)

# Get table of distribution stats for F3 E N group
stats_13 <- data_frame(F3_E_N = gofstat(list(unif$F3_E_N, 
                                             norm$F3_E_N, 
                                             nbin$F3_E_N,
                                             pois$F3_E_N))$aic)

# Get table of distribution stats for F3 E R group (and add an additional row with "NA" to reflect failure to calculate poisson statistics)
stats_14 <- rbind(data_frame(F3_E_R = gofstat(list(unif$F3_E_R,
                                                   norm$F3_E_R, 
                                                   nbin$F3_E_R,
                                                   pois$F3_E_R))$aic), 
                  data_frame(F3_E_R = as.numeric(c("NA"))))

# combine all stats into a single table and add a first column with the name of the distribution being tested
stats <- cbind(distribution = c("uniform", "normal", "negative binomial", "poisson"), 
               stats_1, 
               stats_2, 
               stats_3, 
               stats_4,
               stats_5, 
               stats_6, 
               stats_7,
               stats_8, 
               stats_9, 
               stats_10, 
               stats_11, 
               stats_12, 
               stats_13, 
               stats_14)
```