---
ttitle: "Analysis 5: Herbivory analysis"
author: "Joe Gunn"
date: "2024-09-19"
output: html_document
---

# <b>Project:</b> Assessing phenotypic response and transgenerational inheritance following insecticide exposure in Colorado Potato Beetles (<i>Leptinotarsa decemlineata</i>; CPB) 
<font size="+1"> We assessed inter- and trans-generational effects of sublethal insecticide exposure on four larval performance phenotypes (fecundity, herbivory, mobility, and development) in Colorado potato beetles (<i>Leptinotarsa decemlineata</i>; CPB). We implemented a replicated inbred full-sibling pedigree design, beginning with 12 "founder" mated pair lineages and continuing for four successive generations (full siblings within lineages were inbred in each generation). For the transgenerational experiment, F<sub>0</sub> larvae (offspring of founder pairs) were divided into two "background" groups (exposed, i.e., exposed to a sub-lethal [10-ppm] dose of the neonicotinoid insecticide imidacloprid; and control, i.e., exposed to an equivalent dose of molecular-grade water). All mated pair beetles across lineages, treatments, and generations were monitored for daily fecundity. A subset of third instar larvae were collected from each treatment within each lineage for each generation and monitored for three  performance traits, including larval movement (mobility), feeding rate (herbivory), and time to discrete life stages (development). Performance traits were monitored in three successive generations (F<sub>1</sub>-F<sub>3</sub>) without treatment exposure and compared to the F<sub>0</sub> generation to test for inter- and transgenerationally-inherited "memory" of phenotypic variation. An additional subset of larvae from each generation were re-exposed to the treatments, both experimental (insecticide-exposed) and control (water), to test for potential inheritance of a phenotypic "response" to insecticide stress.</font>

# <b>Specific Aim:</b> Herbivory analysis
<font size="+1">In this analysis, we assess differences in larval herbivory (i.e., total leaf tissue consumed) among mated pairs in the transgenerational experiment by generation (i.e., F0, F1, F2, F3), lineage (i.e., L1, L2, ... L10), and treatment (i.e., control and exposed). </font>

## Phases of Analysis
### Phase 1: Data summary statistics
### Phase 2: Herbivory analysis

### Libraries needed for analysis:
```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(writexl)
library(rstatix)
library(lme4)
library(rsq)
library(scales)
library(lmerTest)
library(fitdistrplus)
library(forcats)
library(patchwork)
library(glmmTMB)
```

## PHASE 1: DATA SUMMARY STATISTICS
In this phase of the analysis, we calculate mean, standard deviation, and variance for larval herbivory metrics within experimental groups of interest.

### STEP 1: Calculate summary statistics for larval herbivory data; run the Rmd chunk below.
In this step, we calculate mean, standard deviation, and variance for larval herbivory (i.e., probability of herbivory and total leaf area consumed) for all experimental groups.

##### Calculate summary statistics for larval herbivory data:
```{r}
# load herbivory data
load("data/herbivory.Rda")

# get sample size (n), mean or proportion, standard deviation (sd), and standard error (se) for herbivory by generation, background, and treatment
herbivory %>%
  group_by(larval_generation, background, treatment) %>%
  summarize(n = n(),
            prop_consumed = sum(consumed)/n,
            sd_consumed = sd(consumed),
            se_consumed = sd_consumed/sqrt(n),
            mean_leaf = mean(leaf_area_consumed),
            sd_leaf = sd(leaf_area_consumed),
            se_leaf = sd_leaf/sqrt(n))
```

These data are presented in Table 1c of the final ms.

## PHASE 2: HERBIVORY ANALYSIS
In this phase of the analysis, we assess differences in larval herbivory (i.e., total leaf area consumed) between treatments and background exposure statuses across generations, accounting for variation in lineages as a random effect in a general linear mixed-effect model framework. We assess two distinct modes of potential phenotypic effects: 1) assessing inheritance of phenotypic "memory" of stress by comparing phenotypic variation (i.e., herbivory metrics) in beetles with a background (F0 exposure) of insecticide vs. those with a background of the water control and without subsequent exposure in successive generations; and 2) assessing inheritance of a phenotypic "response" to stress by comparing phenotypic variation (i.e., herbivory metrics) in beetles treated in each generation with insecticide or a water control and which had a background (F0 exposure) of either insecticide or the water control. For each of these modes of phenotypic effect, we employ general linear mixed effect zero-inflation models to assess simultaneously the probability of herbivory, i.e., the probability that an individual larva will consume leaf tissue and average herbivory, i.e., total leaf consumed.

### STEP 1: Assess phenotypic effects of insecticide stress in total herbivory (leaf area consumed).
In this step, we assess average total herbivory dependent on background exposure and treatment across generations. We first test the fit of raw data to different distributions...

#### 1a: Test fit of empirical herbivory data (leaf area consumed) to a normal distribution; run the Rmd chunk below.
In this step, we use the shapiro-wilk test to assess whether total leaf area consumed among larvae in each group of interest deviates from a normal distribution.

##### Test fit of normality:
```{r}
# load herbivory data
load("data/herbivory.Rda")

# set to show decimals rather than scientific notation for p-value assessments
options(scipen = 999)

# test for fit of normality with treatments and generations and compare to alpha of 0.05/14 = 0.003 (to account for multiple comparisons)
herbivory %>%
  group_by(larval_generation, background, treatment) %>%
  shapiro_test(leaf_area_consumed) %>%
  mutate(alpha = 0.003) %>%
  mutate(deviates = ifelse(p < alpha, "yes", "no"))
```

<b>Analysis results:</b> <br>

<b>Shapiro-wilk p-values for each treatment and generation:</b><br>
F0_C_N: <b>p < 0.001</b><br>
F0_E_N: <b>p < 0.001</b><br>
F1_C_N: p = 0.015 <br>
F1_C_R: <b>p < 0.001</b><br>
F1_E_N: p = 0.005 <br>
F1_E_R: <b>p < 0.001</b><br>
F2_C_N: p = 0.209 <br>
F2_C_R: <b>p < 0.001</b><br>
F2_E_N: p = 0.007 <br>
F2_E_R: <b>p < 0.001</b><br>
F3_C_N: <b>p < 0.001</b><br>
F3_C_R: <b>p < 0.001</b><br>
F3_E_N: <b>p < 0.001</b><br>
F3_E_R: <b>p < 0.001</b><br>

Since 10/14 tests indicate a deviation from a normal distribution (p < 0.003), we test the fit of multiple alternative distributions to determine whether there is consistent best-fitting distribution for the data across groups of interest.

### STEP 2: Test fit of empirical data to alternative distributions.
In this step, we plot the raw data distribution and test the fit of five potential distributions that commonly underlie continuous data (excluding negative poisson, binomial, and negative binomial distributions, given that they are typically used to describe count data), and determine the best fitting distribution for each group of interest. We specifically test uniform, normal, Gamma, exponential, and beta distributions.

#### 2a: Test fit of alternative distributions to the empirical data; run the Rmd chunk below.

##### Test fit of alternative distributions:
```{r}
# load herbivory data
load("data/herbivory.Rda")

# Generate combined "group" factor, which includes larval generation, background, and treatment, to test distributions within each group of interest.
herbivory <- herbivory %>%
  unite(group, larval_generation, c(background, treatment), sep = "_", remove = F) %>%
  mutate(group = factor(group))

# generate list of levels within group factor
levels <- levels(herbivory$group)

# define empty lists to hold statistics for each distribution test
unif <- list() 
norm <- list()
gamm <- list() 
expo <- list()
beta <- list()

# Write for loop to fit each distribution to data, iterating through group levels (i)
for(i in levels) {
  
  # generate subsets of data for each group level
  subset <- herbivory[herbivory$group == i, ]
  
  # fit distributions to leaf area for each subset
  fit_unif <- fitdist(subset$leaf_area_consumed, "unif") 
  fit_norm <- fitdist(subset$leaf_area_consumed, "norm") 
  fit_gamm <- fitdist(subset$leaf_area_consumed, "gamma") 
  fit_expo <- fitdist(subset$leaf_area_consumed, "exp")
  fit_beta <- fitdist(subset$leaf_area_consumed, "beta")
  
  # fill list with fit results for each distribution
  unif[[i]] <- fit_unif 
  norm[[i]] <- fit_norm 
  gamm[[i]] <- fit_gamm 
  expo[[i]] <- fit_expo
  beta[[i]] <- fit_beta
  
}

# Get table of distribution stats for F0 C N group
stats_1 <- data_frame(F0_C_N = gofstat(list(unif$F0_C_N, 
                                            norm$F0_C_N, 
                                            gamm$F0_C_N,
                                            expo$F0_C_N,
                                            beta$F0_C_N))$aic)

# Get table of distribution stats for F0 E N group
stats_2 <- data_frame(F0_E_N = gofstat(list(unif$F0_E_N, 
                                            norm$F0_E_N, 
                                            gamm$F0_E_N,
                                            expo$F0_E_N,
                                            beta$F0_E_N))$aic)

# Get table of distribution stats for F1 C N group
stats_3 <- data_frame(F1_C_N = gofstat(list(unif$F1_C_N, 
                                            norm$F1_C_N, 
                                            gamm$F1_C_N,
                                            expo$F1_C_N,
                                            beta$F1_C_N))$aic)

# Get table of distribution stats for F1 C R group
stats_4 <- data_frame(F1_C_R = gofstat(list(unif$F1_C_R, 
                                            norm$F1_C_R, 
                                            gamm$F1_C_R,
                                            expo$F1_C_R,
                                            beta$F1_C_R))$aic)
         
# Get table of distribution stats for F1 E N group
stats_5 <- data_frame(F1_E_N = gofstat(list(unif$F1_E_N, 
                                            norm$F1_E_N, 
                                            gamm$F1_E_N,
                                            expo$F1_E_N,
                                            beta$F1_E_N))$aic)

# Get table of distribution stats for F1 E R group
stats_6 <- data_frame(F1_E_R = gofstat(list(unif$F1_E_R, 
                                            norm$F1_E_R, 
                                            gamm$F1_E_R,
                                            expo$F1_E_R,
                                            beta$F1_E_R))$aic)

# Get table of distribution stats for F2 C N group
stats_7 <- data_frame(F2_C_N = gofstat(list(unif$F2_C_N, 
                                            norm$F2_C_N, 
                                            gamm$F2_C_N,
                                            expo$F2_C_N,
                                            beta$F2_C_N))$aic)

# Get table of distribution stats for F2 C R group
stats_8 <- data_frame(F2_C_R = gofstat(list(unif$F2_C_R, 
                                            norm$F2_C_R, 
                                            gamm$F2_C_R,
                                            expo$F2_C_R,
                                            beta$F2_C_R))$aic)

# Get table of distribution stats for F2 E N group
stats_9 <- data_frame(F2_E_N = gofstat(list(unif$F2_E_N, 
                                            norm$F2_E_N, 
                                            gamm$F2_E_N,
                                            expo$F2_E_N,
                                            beta$F2_E_N))$aic)

# Get table of distribution stats for F2 E R group
stats_10 <- data_frame(F2_E_R = gofstat(list(unif$F2_E_R, 
                                             norm$F2_E_R, 
                                             gamm$F2_E_R,
                                             expo$F2_E_R,
                                             beta$F2_E_R))$aic)

# Get table of distribution stats for F3 C N group
stats_11 <- data_frame(F3_C_N = gofstat(list(unif$F3_C_N, 
                                             norm$F3_C_N, 
                                             gamm$F3_C_N,
                                             expo$F3_C_N,
                                             beta$F3_C_N))$aic)

# Get table of distribution stats for F3 C R group
stats_12 <- data_frame(F3_C_R = gofstat(list(unif$F3_C_R, 
                                             norm$F3_C_R, 
                                             gamm$F3_C_R,
                                             expo$F3_C_R,
                                             beta$F3_C_R))$aic)

# Get table of distribution stats for F3 E N group
stats_13 <- data_frame(F3_E_N = gofstat(list(unif$F3_E_N, 
                                             norm$F3_E_N, 
                                             gamm$F3_E_N,
                                             expo$F3_E_N,
                                             beta$F3_E_N))$aic)

# Get table of distribution stats for F3 E R group
stats_14 <- data_frame(F3_E_R = gofstat(list(unif$F3_E_R,
                                             norm$F3_E_R, 
                                             gamm$F3_E_R,
                                             expo$F3_E_R,
                                             beta$F3_E_R)))

# Run global distribution to test if gamma is best fitting

# Run individual global distributions 
unif <- fitdist(herbivory$leaf_area_consumed, "unif")
norm <- fitdist(herbivory$leaf_area_consumed, "norm")
gamm <- fitdist(herbivory$leaf_area_consumed, "gamma")
expo <- fitdist(herbivory$leaf_area_consumed, "exp")
beta <- fitdist(herbivory$leaf_area_consumed, "beta")

# compare AIC values 
gofstat(list(unif, 
             norm,
             gamm, 
             expo, 
             beta),
        fitnames = c("uniform",
                     "normal",
                     "gamma",
                     "exponential",
                     "beta"))$aic
```

<b>Analysis results:</b> <br>

| Group     | Uniform          | Normal           | Gamma             | Exponential        | Beta              |
|:----------|:-----------------|:-----------------|:----------------- | :----------------- |:----------------- |
| F0_C_N    | -223.503         | -205.761         | <b>-88.812</b>    | -323.386           | -89.887           |
| F0_E_N    | -203.925         | -197.072         | <b>-69.398</b>    | -332.413	         | -70.716           |
| F1_C_N    | -85.261          | -80.147          | <b>-66.657</b>    | -100.362           | -67.11998         |
| F1_C_R    | -107.474         | -120.673         | <b>-16.505</b>    | -223.128           | -16.573           |
| F1_E_N    | -75.030          | -63.358          | <b>-59.169</b>    | -85.599            | -59.426           |
| F1_E_R    | -128.401         | -121.989         | <b>-17.686</b>    | -198.561           | -17.770	         |
| F2_C_N    | -55.271          | -48.551          | <b>-46.886</b>    | -54.286            | -47.167           |
| F2_C_R    | -80.299          | -83.199          | <b>-19.301</b>    | -147.651           | -19.397           |
| F2_E_N    | -50.383          | -40.278          | <b>-26.729</b>    | -60.092            | -27.258           |
| F2_E_R    | -79.023          | -71.588          | <b>-23.297</b>    | -108.993           | -23.447           |
| F3_C_N    | -40.748          | -33.974          | <b>-17.740</b>    | -62.307            | -17.80904         |
| F3_C_R    | -57.926          | -61.074          | <b>-5.465</b>     | -127.7364          | -5.465403         |
| F3_E_N    | -74.813          | -73.264          | <b>-20.262</b>    | -126.413           | -20.312           |
| F3_E_R    | -106.766         | -114.668         | <b>-5.106</b>     | -205.987           | -5.159            |
| Global    | -1122.154        | -1159.560        | <b>-487.994</b>   | -1909.659          | -494.259          |

The data in all groups were best fit to a Gamma distribution, including the global dataset (all groups aggregated). We therefore opted to assume a Gamma distribution in all general linear model analyses.

These data are presented as Table XXX in the final ms.

#### 2b: Plot empirical data distribution; run the Rmd chunk below.

##### Plot empirical data distribution for leaf area consumed: `figures/distribution.jpeg`
```{r}
# load herbivory data
load("data/herbivory.Rda")

# plot distribution
jpeg("figures/distribution.jpeg", width = 5000, height = 4000, res = 500)

ggplot(herbivory, aes(x = leaf_area_consumed)) +
  geom_histogram(binwidth = 0.005, color = "black", fill = "grey", linewidth = 1) +
  theme_set(theme_cowplot(12)) +
  scale_x_continuous(breaks = seq(from = -0.05, to = 0.3, by = 0.05),
                     limits = c(-0.05,0.3)) +
  scale_y_continuous(breaks = seq(from = 0, to = 270, by = 30),
                     limits = c(0,270)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 30, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  xlab(expression("Leaf area consumed"~(cm^{"2"}))) +
  ylab("Frequency") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

dev.off()
```

This figure is presented as Figure XXX in the final ms.

### STEP 3: Assess phenotypic effects of insecticide stress in total herbivory walked.

#### 4a: Run two-factor general linear mixed effect interaction models assuming a zero-inflated gamma distribution to test the relative effects of background and treatment of insecticide stress on herbivory; run the Rmd chunk below.
In this step, we run general linear mixed effect interaction models with raw herbivory (leaf area consumed) as a response, interacting fixed effects of background and treatment, and lineage as a random effect. We assume an underlying Gamma distribution for the data (see Step 3 above). We also assume zero-inflation of the data, given that there may be distinct processes driving the ability of an individual larva to eat at all (zero or non-zero) and how much they are able to eat if they can (non-zero). We therefore run a zero-inflation gamma model, estimating coefficients of both the zero/non-zero model (i.e., zero-inflation; logistic regression) and the conditional model (non-zero; general linear regression).

##### Run general linear mixed effect interaction models for herbivory:
```{r}
# load herbivory data
load("data/herbivory.Rda")

# set to show decimals rather than scientific notation for p-value assessments
options(scipen = 999)

# isolate only F0 data
f0 <- herbivory %>%
  filter(larval_generation == "F0")

# isolate only F1 data
f1 <- herbivory %>%
  filter(larval_generation == "F1")

# isolate only F0 data
f2 <- herbivory %>%
  filter(larval_generation == "F2")

# isolate only F0 data
f3 <- herbivory %>%
  filter(larval_generation == "F3")

# f0 general linear mixed effect zero-inflation model with conditional beta distribution
f0_mod <- glmmTMB(leaf_area_consumed ~ background + (1|lineage),
                  data = f0,
                  ziformula = ~background,
                  family = ziGamma(link = "log"))

# f1 general linear mixed effect zero-inflation model with conditional beta distribution
f1_mod <- glmmTMB(leaf_area_consumed ~ treatment*background + (1|lineage),
                  data = f1,
                  ziformula = ~treatment*background,
                  family = ziGamma(link = "log"))

# f2 general linear mixed effect zero-inflation model with conditional beta distribution
f2_mod <- glmmTMB(leaf_area_consumed ~ treatment*background + (1|lineage),
                  data = f2,
                  ziformula = ~treatment*background,
                  family = ziGamma(link = "log"))

# f2 general linear mixed effect zero-inflation model with conditional beta distribution
f3_mod <- glmmTMB(leaf_area_consumed ~ treatment*background + (1|lineage),
                  data = f3,
                  ziformula = ~treatment*background,
                  family = ziGamma(link = "log"))

# summarize f0 model
summary(f0_mod) 
df.residual(f0_mod)

# summarize f1 model
summary(f1_mod) 
df.residual(f1_mod)

# summarize f2 model
summary(f2_mod) 
df.residual(f2_mod)

# summarize f3 model
summary(f3_mod) 
df.residual(f3_mod)
```

<b>Analysis results:</b> <br>

<b>Coefficients of zero-inflated logistic regressions by generation:</b><br>
<b>F0</b>: <br>
Background: z = 0.679; p = 0.497 <br>

<b>F1</b>: <br>
Treatment: z = 4.171; p < 0.001 <br>
Background: z = -0.468; p = 0.6401 <br>
Treatment x Background : z = 0.266; p = 0.7901 <br>

<b>F2</b>: <br>
Treatment: z = 3.493; p = 0.000477 <br>
Background: z = 1.197; p = 0.231469 <br>
Treatment x Background: z = -1.303; p = 0.192407 <br>

<b>F3</b>: <br>
Treatment: z = 2.083; p = 0.0373 <br>
Background: z = 1.316; p = 0.1881 <br>
Treatment x Background: z = -0.229; p = 0.8186 <br>

b>Coefficients of conditional general linear regressions by generation:</b><br>
<b>F0</b>: <br>
Background: z = 0.35; p = 0.726 <br>

<b>F1</b>: <br>
Treatment: z = -1.645; p = 0.0999 <br>
Background: z = 0.335; p = 0.7376 <br>
Treatment x Background : z = 0.357; p = 0.7214 <br>

<b>F2</b>: <br>
Treatment: z = -1.993; p = 0.0463 <br>
Background: z = -0.923; p = 0.3562 <br>
Treatment x Background: z = -0.002; p = 0.9985 <br>

<b>F3</b>: <br>
Treatment: z = -0.336; p = 0.737 <br>
Background: z = -1.536; p = 0.125 <br>
Treatment x Background: z = 0.385; p = 0.700 <br>

These data are presented in Table 2b in the final ms.

### STEP 3: Plot mobility data analyses

#### 3a: Plot probability of hervivory data; run the Rmd chunk below.

##### Plot probability data: `figures/herbivory_probability.jpeg`
```{r}
# load herbivory data
load("data/herbivory.Rda")

# get sample size (n), mean or proportion, standard deviation (sd), and standard error (se) for herbivory by generation, background, and treatment for plotting
herbivory <- herbivory %>%
  group_by(larval_generation, background, treatment) %>%
  summarize(n = n(),
            prop_consumed = sum(consumed)/n,
            sd_consumed = sd(consumed),
            se_consumed = sd_consumed/sqrt(n)) %>%
  mutate(background = recode_factor(background, C = "Control", E = "Insecticide"),
         treatment = recode_factor(treatment, N = "Not re-exposed", R = "Re-exposed"))

# isolate only F0 data
f0 <- herbivory %>%
  filter(larval_generation == "F0")

# isolate only F1 data
f1 <- herbivory %>%
  filter(larval_generation == "F1")

# isolate only F2 data
f2 <- herbivory %>%
  filter(larval_generation == "F2")

# isolate only F3 data
f3 <- herbivory %>%
  filter(larval_generation == "F3")

# plot f0 data
f0_plot <- ggplot(f0, aes(x = background, y = prop_consumed, group = background, color = background)) +
  geom_errorbar(aes(ymin = prop_consumed - se_consumed, ymax = prop_consumed + se_consumed, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.2, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = background, y = prop_consumed, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 7, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = F) +
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.2)) +
  ylim(0,1) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 30, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  labs(x = "Original exposure", y = "P(Herbivory)") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f1 data
f1_plot <- ggplot(f1, aes(x = treatment, y = prop_consumed, group = background, color = background)) +
  geom_line(aes(color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = F) + 
  geom_errorbar(aes(ymin = prop_consumed - se_consumed, ymax = prop_consumed + se_consumed, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.2, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = treatment, y = prop_consumed, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 7, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = F) +
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.2)) +
  ylim(0,1) +
  theme(axis.title = element_text(size = 30)) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title.y = element_blank()) +
  theme(axis.title.x = element_blank()) +
  #theme(legend.text = element_text(size = 30)) +
  #theme(legend.title = element_text(size = 30, face = "bold")) +
  #theme(legend.key.size = unit(2, 'cm')) +
  #theme(legend.key.spacing.y = unit(-0.5, 'cm')) +
  #theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  #theme(legend.box.margin = margin(5, 10, 1, 10)) +
  #theme(legend.position = c(0.05,0.2)) + 
  labs(x = "Treatment", y = "P(Herbivory)") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f2 data
f2_plot <- ggplot(f2, aes(x = treatment, y = prop_consumed, group = background, color = background)) +
  geom_line(aes(color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = F) + 
  geom_errorbar(aes(ymin = prop_consumed - se_consumed, ymax = prop_consumed + se_consumed, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.2, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = treatment, y = prop_consumed, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 7, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = F) +
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.2)) +
  ylim(0,1) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  labs(x = "Treatment", y = "P(Herbivory)") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f3 data
f3_plot <- ggplot(f3, aes(x = treatment, y = prop_consumed, group = background, color = background)) +
  geom_line(aes(color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = F) + 
  geom_errorbar(aes(ymin = prop_consumed - se_consumed, ymax = prop_consumed + se_consumed, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.2, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = treatment, y = prop_consumed, group = background, fill = background),
             position = position_dodge(width = 0.4), 
             size = 7,
             pch = 22,
             stroke = 2, 
             color = "black", 
             show.legend = F) +
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.2)) +
  ylim(0,1) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30)) + 
  theme(axis.title.y = element_blank()) +
  theme(axis.title.x = element_text(vjust = -0.5)) +
  labs(x = "Treatment", y = "P(Herbivory)") +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot all distance figures together
jpeg("figures/herbivory_probability.jpeg", width = 17000, height = 15000, res = 1000)

f0_plot + 
  f1_plot + 
  f2_plot + 
  f3_plot + 
  plot_layout(nrow = 2, ncol = 2)

dev.off()
```

This figure is presented as Figure XXX in the final ms.

#### 3b: Plot means of hervivory data; run the Rmd chunk below.

##### Plot mean data: `figures/herbivory_means.jpeg`
```{r}
# load herbivory data
load("data/herbivory.Rda")

# get sample size (n), mean or proportion, standard deviation (sd), and standard error (se) for herbivory by generation, background, and treatment for plotting
herbivory <- herbivory %>%
  group_by(larval_generation, background, treatment) %>%
  summarize(mean = mean(leaf_area_consumed),
            sd = sd(leaf_area_consumed),
            n = n(),
            se = sd/sqrt(n)) %>%
  mutate(background = recode_factor(background, C = "Control", E = "Insecticide"),
         treatment = recode_factor(treatment, N = "Not re-exposed", R = "Re-exposed"))

# isolate only F0 data
f0 <- herbivory %>%
  filter(larval_generation == "F0")

# isolate only F1 data
f1 <- herbivory %>%
  filter(larval_generation == "F1")

# isolate only F2 data
f2 <- herbivory %>%
  filter(larval_generation == "F2")

# isolate only F3 data
f3 <- herbivory %>%
  filter(larval_generation == "F3")

# plot f0 data
f0_plot <- ggplot(f0, aes(x = background, y = mean, group = background, color = background)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.2, 
                linewidth = 2, 
                show.legend = F) +
   geom_point(aes(x = background, y = mean, group = background, fill = background), 
              position = position_dodge(width = 0.4), 
              size = 7, 
              pch = 22, 
              stroke = 2, 
              color = "black", 
              show.legend = F) +
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 0.15, by = 0.05)) +
  ylim(0,0.15) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(axis.text = element_text(size = 30, hjust = 0.4)) + 
  theme(axis.title = element_text(size = 30, hjust = 0.5)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  labs(x = "Original exposure") +
  ylab(expression("Leaf area consumed"~(cm^{"2"}))) +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f1 data
f1_plot <- ggplot(f1, aes(x = treatment, y = mean, group = background, color = background)) +
  geom_line(aes(color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = F) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.2, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = treatment, y = mean, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 7, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = F) +
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 0.15, by = 0.05)) +
  ylim(0,0.15) +
  theme(axis.title = element_text(size = 30)) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title.y = element_blank()) +
  theme(axis.title.x = element_blank()) +
  #theme(legend.text = element_text(size = 30)) +
  #theme(legend.title = element_text(size = 30, face = "bold")) +
  #theme(legend.key.size = unit(2, 'cm')) +
  #theme(legend.key.spacing.y = unit(-0.5, 'cm')) +
  #theme(legend.box.background = element_rect(color = "black", size = 1.5)) +
  #theme(legend.box.margin = margin(5, 10, 1, 10)) +
  #theme(legend.position = c(0.05,0.2)) + 
  labs(x = "Treatment") +
  ylab(expression("Leaf area consumed"~(cm^{"2"}))) +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f2 data
f2_plot <- ggplot(f2, aes(x = treatment, y = mean, group = background, color = background)) +
  geom_line(aes(color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = F) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.2, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = treatment, y = mean, group = background, fill = background), 
             position = position_dodge(width = 0.4), 
             size = 7, 
             pch = 22, 
             stroke = 2, 
             color = "black", 
             show.legend = F) +
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 0.15, by = 0.05)) +
  ylim(0,0.15) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30)) + 
  theme(axis.title.y = element_text(vjust = +3)) + 
  theme(axis.title.x = element_text(vjust = -0.5)) +
  labs(x = "Treatment") +
  ylab(expression("Leaf area consumed"~(cm^{"2"}))) +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot f3 data
f3_plot <- ggplot(f3, aes(x = treatment, y = mean, group = background, color = background)) +
  geom_line(aes(color = background), 
            position = position_dodge(width = 0.4), 
            linewidth = 2, 
            show.legend = F) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = background), 
                position = position_dodge(width = 0.4), 
                width = 0.2, 
                linewidth = 2, 
                show.legend = F) +
  geom_point(aes(x = treatment, y = mean, group = background, fill = background),
             position = position_dodge(width = 0.4), 
             size = 7,
             pch = 22,
             stroke = 2, 
             color = "black", 
             show.legend = F) +
  theme_set(theme_cowplot(12)) +
  scale_color_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_fill_manual(values = c("deepskyblue2", "firebrick3")) +
  scale_y_continuous(breaks = seq(from = 0, to = 0.15, by = 0.05)) +
  ylim(0,0.15) +
  theme(axis.text = element_text(size = 30)) + 
  theme(axis.title = element_text(size = 30)) + 
  theme(axis.title.y = element_blank()) +
  theme(axis.title.x = element_text(vjust = -0.5)) +
  labs(x = "Treatment") +
  ylab(expression("Leaf area consumed"~(cm^{"2"}))) +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2)) +
  theme(plot.margin = margin(20,20,20,20))

# plot all herbivory figures together
jpeg("figures/herbivory_means.jpeg", width = 17000, height = 15000, res = 1000)

f0_plot + 
  f1_plot + 
  f2_plot + 
  f3_plot + 
  plot_layout(nrow = 2, ncol = 2)

dev.off()
```

This figure is presented as Figure XXX in the final ms.
